\section{Proposed Method}
In this section, we present the architecture and theoretical formulation of the Cross-Modality Semantic Alignment Transformer (CMSA-Trans). The core objective of the proposed framework is to bridge the gap between high-dimensional vibration signals and abstract technical concepts by using the extensive mechanical knowledge built-in to large language models (LLMs). By mapping physical signal features into a shared semantic space, the model enables zero-shot fault diagnosis of rotating machinery. In this approach, health states of unseen components are recognized without prior training samples.

Rotating machinery, such as induction motors, wind turbines, and industrial gearboxes, forms the backbone of modern manufacturing. However, these systems are prone to various failure modes due to harsh operational environments and long-term fatigue. Traditional supervised learning methods perform well when labeled data is abundant for all conditions, but they falter in zero-shot scenarios where certain failure modes have never been recorded. The proposed CMSA-Trans addresses this by treating fault diagnosis not as a simple classification task, but as a cross-modal retrieval problem. By embedding both signals and linguistic descriptions into a common manifold, inference is performed on unseen categories by providing textual descriptions to the LLM.

\subsection{Overall Architecture}
The overall architecture of the proposed CMSA-Trans is illustrated in Fig.~\ref{fig:framework}. The framework is designed to transform raw sensor data into semantic-aligned representations that are directly comparable to linguistic descriptions of machinery faults. The high-level workflow begins with the acquisition of raw vibration signals from sensors mounted on rotating components, such as bearings or gearboxes. These signals are typically non-stationary and contain periodic impulsive components that signify specific fault modes.

The architecture consists of four primary functional modules: a signal encoder, a semantic prototyping module, a cross-modality alignment layer, and a zero-shot classification head. The signal encoder uses a transformer-based structure \cite{vaswani2017attention, zerveas2021transformer} to extract deep temporal features from the vibration data. At the same time, the semantic prototyping module interfaces with a pre-trained LLM \cite{radford2021learning} to generate high-fidelity technical descriptions for each fault category. These descriptions are then converted into fixed-dimensional semantic vectors (prototypes). This automation removes the need for expert-defined attribute matrices.

The heart of the CMSA-Trans is the cross-modality alignment module. This module employs a mutual-attention mechanism to project the learned signal features into the semantic space. Unlike traditional diagnostic models that map signals directly to discrete class labels, the proposed approach maps signals to a continuous semantic space defined by the LLM. This allows the model to reason about the similarity between a physical signal pattern and a technical concept (e.g., ``outer race pitting''). This alignment process is inspired by recent advances in contrastive learning architectures like CLIP \cite{radford2021learning} and cross-modal projection frameworks \cite{oord2018representation, chen2020simple}. For instance, a signal characterized by sharp transients at specific intervals is matched with a textual description that defines those intervals as the characteristic ball pass frequency. Finally, zero-shot classification is performed by identifying the nearest semantic prototype for a given signal embedding. This structured approach ensures that the model captures the underlying physics of machine failures rather than just memorizing data patterns. This method helps generalization to new, unseen fault modes. By using the semantic bridge provided by the LLM, CMSA-Trans effectively enables zero-shot diagnosis for industrial applications.

\subsection{Signal Encoder: Time-Series Transformer}
Vibration signals from rotating machinery are characterized by multi-scale periodicity and non-stationary impulses. Traditional deep learning backbones, such as convolutional neural networks (CNNs), often fail to capture the long-range dependencies between impacts occurring across multiple shaft revolutions due to local receptive fields. While recurrent neural networks (RNNs) can handle sequential data, they suffer from vanishing gradients when processing long industrial signal sequences. To address these limitations, a time-series transformer (TST) is used as the signal encoder \cite{zerveas2021transformer}. This encoder uses self-attention to maintain a global view of the entire signal window.

Given a raw vibration signal segment $\mathbf{X} \in \mathbb{R}^{L \times 1}$, where $L$ is the window length, the signal is first partitioned into $N$ non-overlapping patches $\mathbf{x}_p \in \mathbb{R}^{P \times 1}$, with $P$ denoting the patch size. Each patch represents a localized temporal window of the vibration waveform. These patches are then projected into a high-dimensional embedding space using a linear layer with weights $\mathbf{W}_b \in \mathbb{R}^{P \times d_m}$, where $d_m$ is the model dimension. This projection allows the model to process 1D temporal segments as discrete tokens. To preserve the relative and absolute temporal order of the signal, learnable positional encodings $\mathbf{E}_{pos} \in \mathbb{R}^{N \times d_m}$ are added to the projected embeddings. The resulting sequence of input embeddings $\mathbf{Z}_0$ is formulated as:
\begin{equation}
\mathbf{Z}_0 = [\mathbf{x}_p^{(1)}\mathbf{W}_b; \mathbf{x}_p^{(2)}\mathbf{W}_b; \dots; \mathbf{x}_p^{(N)}\mathbf{W}_b] + \mathbf{E}_{pos}
\end{equation}
where $\mathbf{Z}_0 \in \mathbb{R}^{N \times d_m}$ serves as the input to the transformer backbone.

The embedded sequence passes through a stack of transformer encoder layers. Each layer consists of a multi-head self-attention (MHSA) sub-layer and a position-wise feed-forward network (FFN). The MHSA mechanism is particularly effective for vibration signals. It allows each segment of the signal to attend to every other segment, regardless of the distance in time. This is key for detecting fault signatures like inner-race faults, where the impulsive pattern is modulated by the shaft rotation and may be separated by long intervals of background noise. The FFN layer then applies non-linear transformations to each token independently, refining the feature representation. Residual connections and layer normalization are used to ensure stable training. The output of the final transformer layer, denoted as $\mathbf{H}_v \in \mathbb{R}^{N \times d_m}$, represents the refined long-range structural features. This encoder effectively reduces the inductive bias of local-receptive field models and enables reliable feature extraction in the presence of industrial background noise. Through this architecture, the TST captures the rhythm of the machine, which is fundamental for subsequent semantic alignment.

\subsection{Semantic Prototyping via LLM}
A major bottleneck in zero-shot fault diagnosis is the construction of a discriminative semantic space. Conventional methods rely on manual attribute engineering (e.g., binary labels for ``impulsive'' or ``harmonic''). Such work is labor-intensive, subjective, and fails to scale to complex industrial systems with hundreds of failure modes. The proposed CMSA-Trans overcomes this by automating semantic generation through an LLM interface, treating the LLM as a virtual domain expert.

The world knowledge of a pre-trained LLM (e.g., GPT-4) is used to generate detailed technical descriptions for each fault mode. Specifically, for each seen class in the training set and unseen class in the test set, a structured prompt is provided: ``Provide a detailed technical description of the vibration characteristics for a [Fault Name] in a bearing, focusing on its frequency components, impact patterns, and resonance bands.'' The LLM generates a text corpus that describes the physical manifestations of the fault, effectively linking the label to mechanical engineering principles. This process is better than manual labeling because the LLM synthesizes information from a large array of technical manuals and research papers.

Table~\ref{tab:llm_semantics} presents a comparison of the LLM-generated semantic descriptions with traditional word-level embeddings. These textual descriptions are then transformed into fixed-length semantic vectors (prototypes) using a sentence-level embedding model such as SBERT \cite{reimers2019sentence}. For a class $y$, the LLM-generated text $T_y$ is encoded into a semantic prototype $\mathbf{a}_y \in \mathbb{R}^{d_s}$, where $d_s$ is the dimensionality of the semantic space. These prototypes $\mathbf{a}_y$ act as anchors in the embedding space. Since the LLM uses external technical knowledge to define these anchors, the resulting semantic space is grounded in the physics of failures. This allows the model to distinguish between similar fault types based on small differences in technical descriptions. These include the specific ball pass frequency (BPF) characteristics or sideband patterns mentioned in the text. Also, because these prototypes are generated in a continuous vector space, the model represents the relationships between faults. For instance, two different seal-rubbing faults naturally have similar semantic vectors. This rich semantic structure enables the cross-modal alignment module to generalize signal features to previously unobserved conditions.

\subsection{Cross-Modality Alignment}
The core innovation of the proposed method is the cross-modality alignment module. This module maps the vibration signal features $\mathbf{H}_v$ into the semantic space $\mathcal{S}$ populated by the LLM-derived prototypes. This mapping is not a simple linear projection but a dynamic process that identifies which physical signal features correspond to the technical descriptors. This stage is vital for overcoming the modality gap, where raw numerical data and linguistic concepts reside in different distributions.

A mutual-attention mechanism is used where the signal features $\mathbf{H}_v$ act as the query, and the semantic prototypes of the seen classes $\mathbf{A}_{seen} \in \mathbb{R}^{C_{seen} \times d_s}$ act as the keys and values. This allows the model to learn a cross-modal sensitivity. When the LLM description mentions a ``high-frequency impact,'' the attention mechanism puts more weight on the spectral tokens within $\mathbf{H}_v$ that exhibit such impulsiveness. This dynamic selection process ensures that the model focuses on diagnostic-relevant information rather than background noise. The query ($\mathbf{Q}$), key ($\mathbf{K}$), and value ($\mathbf{V}$) matrices for the alignment layer are defined as:
\begin{equation}
\mathbf{Q} = \text{Norm}(\mathbf{H}_v) \mathbf{W}_Q, \quad \mathbf{K} = \mathbf{A}_{seen} \mathbf{W}_K, \quad \mathbf{V} = \mathbf{A}_{seen} \mathbf{W}_V
\end{equation}
where $\mathbf{W}_Q$, $\mathbf{W}_K$, and $\mathbf{W}_V$ are learnable weight matrices. The cross-modality attention score is then computed to aggregate the signal features aligned with the semantic context:
\begin{equation}
\mathbf{H}_{aligned} = \text{softmax}\left(\frac{\mathbf{Q}\mathbf{K}^T}{\sqrt{d_k}}\right) \mathbf{V}
\end{equation}
where $\mathbf{H}_{aligned}$ represents the semantic-aware signal representation and $d_k$ is the scaling factor. This ensures that the alignment is driven by the intrinsic relationship between signal morphology and technical terminology.

Following the attention layer, global average pooling is applied to the aligned sequence, which is then passed through a final projection layer $\mathbf{W}_p \in \mathbb{R}^{d_m \times d_s}$. The final signal embedding $\mathbf{v}_x \in \mathbb{R}^{d_s}$ in the semantic space is given by:
\begin{equation}
\mathbf{v}_x = \text{ReLU}(\text{Pooling}(\mathbf{H}_{aligned}) \mathbf{W}_p)
\end{equation}
By mapping $\mathbf{v}_x$ into the same $d_s$-dimensional space as the LLM prototypes, a unified platform is created for direct similarity measurement between signals and textual concepts. This projection layer $\mathbf{W}_p$ is basically a translation matrix that converts structural vibration features into the semantic dialect of the LLM.

\subsection{Training and Optimization}
The training phase aims to align the signal embedding $\mathbf{v}_x$ with its corresponding semantic prototype $\mathbf{a}_{y}$ for seen classes. A contrastive alignment loss \cite{chen2020simple} based on cosine similarity is used to pull the signal embedding towards its correct prototype while pushing it away from all other prototypes in the seen set. This helps the learning of an embedding space where clusters are formed based on health-state semantics. This contrastive approach is strong against intra-class variations, as it focuses on invariant semantic features.

For a training batch of size $B$, the total loss function $\mathcal{L}_{total}$ is defined as a multi-class cross-entropy variant applied to cosine distances:
\begin{equation}
\mathcal{L}_{total} = -\frac{1}{B} \sum_{i=1}^{B} \log \frac{\exp(\text{cos}(\text{v}_x^{(i)}, \mathbf{a}_{y_i}) / \tau)}{\sum_{j \in \mathcal{Y}_{seen}} \exp(\text{cos}(\mathbf{v}_x^{(i)}, \mathbf{a}_j) / \tau)}
\end{equation}
where $\text{cos}(\cdot)$ is the cosine similarity between two vectors, $\tau$ is a temperature hyperparameter, and $\mathcal{Y}_{seen}$ is the set of seen class labels. The temperature parameter controls the sharpness of the probability distribution over classes.

During inference, a test vibration signal $\mathbf{X}_{test}$ is passed through the signal encoder and alignment module to obtain its semantic embedding $\mathbf{v}_{test}$. Zero-shot classification is then performed by identifying the nearest semantic prototype among all candidate classes, including unseen ones. The predicted label $\hat{y}$ is determined by the following utility function:
\begin{equation}
\hat{y} = \arg\max_{c \in \mathcal{Y}_{all}} \text{cos}(\mathbf{v}_{test}, \mathbf{a}_c)
\end{equation}
where $\mathcal{Y}_{all} = \mathcal{Y}_{seen} \cup \mathcal{Y}_{unseen}$ encompasses both seen and unseen fault categories, and $\mathbf{a}_c$ denotes the LLM-generated semantic prototype for class $c$. This formulation enables the model to generalize to new fault types without retraining. The semantic prototypes for unseen classes are generated independently by the LLM.
