\begin{abstract}
Vibration-based fault diagnosis stands as a cornerstone of predictive maintenance in industrial systems, yet traditional methods struggle with data scarcity, domain shifts, and limited generalization across machinery. This paper introduces VibSemAlign, a novel framework that leverages multimodal large language models (MLLMs) to align vibration signals with semantic fault descriptions, enabling robust zero-shot and few-shot diagnosis. By converting raw vibrations into spectrograms and pairing them with textual prompts describing fault types, VibSemAlign fine-tunes vision-language models through a specialized vibration-semantic contrastive loss. This loss fosters deep alignment between visual signal representations and natural language semantics. Integrated with model-agnostic meta-learning (MAML), the framework achieves rapid adaptation to new domains with minimal samples. Extensive experiments on benchmarks like CWRU and Paderborn datasets demonstrate superior performance: up to 15\% accuracy gains in cross-domain zero-shot settings compared to state-of-the-art deep learning baselines. VibSemAlign not only outperforms traditional signal processing and CNN-based classifiers but also provides interpretable semantic reasoning for faults. Key contributions include the first MLLM-based semantic alignment for vibrations, a tailored contrastive objective, and validated generalization across industrial datasets. This work paves the way for semantically grounded, data-efficient fault diagnosis in real-world deployments.
\end{abstract}
