\documentclass[lettersize,journal]{IEEEtran}

\usepackage{amsmath,amsfonts}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{url}
\usepackage{stfloats}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}

\title{VibSemAlign: Vibration-Semantic Alignment Framework for Multimodal Large Model-based Intelligent Fault Diagnosis}

\author{Anonymous}

\maketitle

\begin{abstract}
Vibration-based fault diagnosis remains challenged by domain discrepancies and limited labeled data in industrial deployments. This paper presents VibSemAlign, a framework that aligns vibration spectrograms with semantic fault descriptions via vision-language models (VLMs). VibSemAlign introduces a vibration-semantic contrastive loss (VSCL) for aligning spectrogram and text embeddings in a shared latent space, an 8-head cross-attention decoder for multimodal fusion, and Model-Agnostic Meta-Learning (MAML) for rapid few-shot adaptation to unseen domains. Physics-informed prompt engineering encodes fault characteristic frequencies and operating conditions into natural language descriptors. Evaluated on CWRU and Paderborn bearing datasets, VibSemAlign achieves 96.8\% intra-domain accuracy and 85.3\% zero-shot cross-domain transfer (CWRU$\to$Paderborn), surpassing state-of-the-art methods by 4.7--17.2 percentage points. Ablation studies confirm the orthogonality of contrastive alignment, cross-attention fusion, and meta-learning components. The framework demonstrates that VLMs, when adapted through domain-specific contrastive objectives, serve as effective reasoning engines for semantic-driven industrial diagnostics.
\end{abstract}

\begin{IEEEkeywords}
Fault diagnosis, vision-language model, contrastive learning, vibration spectrogram, zero-shot learning, meta-learning, cross-attention.
\end{IEEEkeywords}

\input{introduction}
\input{related_work}
\input{method}
\input{experiments}
\input{conclusion}

\bibliographystyle{plain}
\bibliography{references}

\end{document}
