\begin{abstract}
Zero-shot fault diagnosis (ZSFD) addresses the critical challenge of identifying unseen fault categories in rotating machinery where labeled data is completely missing. While generative models, particularly diffusion models, have recently demonstrated remarkable capabilities in synthesizing virtual fault samples, they predominantly operate as data-driven black boxes. Consequently, the generated signals often violate fundamental mechanical dynamics and lack physical reliability. Furthermore, current ZSFD frameworks fail to quantify diagnostic uncertainty, limiting their trustworthiness in safety-critical industrial applications. To bridge these gaps, this paper proposes a novel Physics-Constrained Diffusion Model (PC-Diffusion) coupled with a Bayesian semantic embedding framework. Initially, a Physics-Informed Neural Network (PINN) is employed to embed the differential equations of bearing vibrations, extracting a physics-informed prior. Subsequently, PC-Diffusion generates virtual samples guided by physical constraints, specifically frequency-domain energy conservation and time-domain periodicity mechanisms. A Bayesian semantic embedding network further maps dual-level semantics—combining manual physical attributes and high-dimensional embeddings from Large Language Models (LLMs)—into the feature space via variational inference, enabling robust uncertainty quantification. Extensive experiments on the CWRU and XJTU-SY datasets demonstrate that the proposed method not only achieves state-of-the-art diagnostic accuracy on unseen faults but also provides reliable confidence intervals, significantly enhancing the interpretability and reliability of generative zero-shot diagnosis.
\end{abstract}